XRONOS: An Open Data Infrastructure for Archaeological Chronology

### 2025-04-04 paper.qmd
Update in response to reviews

Datasets based on literature review also become out of date almost immediately upon publication, due the the constant production of new dates.
Unfortunately this applies to many databases that are in theory continuously updated, as it is common to see them become unmaintained and or unexpectedly become unavailable.
Of the `r n_c14_datasets` published datasets we identified, `r n_c14_datasets_maintained` were intended to be continuously updated, but only `r n_c14_datasets_recently_updated` have received updates in the last two years.
The average 'lifespan' of a dataset from its publication to its last update is around `r english(round(mean_c14_datasets_lifespan))` years.
Most radiocarbon datasets we reviewed were compiled with a specific goal in mind (e.g. a particular analysis) and, even where there is the intention to keep them updated afterwards, the exigencies of scientific production combined with the labour-intensive nature of the process make that difficult to achieve in practice.

@@

```{r fig-c14-global, message=FALSE, warning=FALSE}
#| fig-cap: Geographic (kernel density) and temporal (sum calibration) of georeferenced dates in XRONOS and other global radiocarbon compilations
#| fig-height: 5
plot_c14_density <- function(density, title, coastline = coast) {
  ggplot() +
    geom_stars(
      aes(fill = ecdf(v)(v)), # transform density to quantiles like plot.stars()
      data = density
    ) +
    geom_sf(data = coastline, fill = NA) +
    scale_fill_stepsn(
      name = "Density",
      n.breaks = 11,
      colours = colour("turku", reverse = TRUE)(11),
      labels = c("Low", rep("", 7), "High"),
      na.value = NA,
      guide = guide_none()
    ) +
    labs(title = title) +
    coord_sf(crs = "+proj=moll") +
    theme_minimal_grid(font_size = 11) +
    theme(
      plot.title = element_text(hjust = 0.5),
      plot.margin = unit(c(11, 5.5, 0, 5.5), "pt")
    )
}

@@

Radiocarbon has been by far the most active area of open data compilation, but archaeological chronology incorporates a much more diverse range of sources of information [@Harding1999].
In periods beyond the practical limit of radiocarbon dating (c. 55,000 BP), other types of radiometric (K–Ar, U–Pb, etc.), chemical or luminescence dating offer an alternative [@Aitken1999].
Conversely, in historic periods, or on calibration 'plateaus' [@GuldersonEtAl2005], radiocarbon is often less important than conventional typological dating (based on artefact characteristics), which in these periods can offer comparable or better temporal resolution, or direct dating based on epigraphy [@HermankovaEtAl2021], numismatics [@KemmersMyrberg2011] or historical sources.
In places where it is widely available, dendrochronology [@Baillie2014] also produces significantly better resolved chronologies and therefore tends to be the main source of chronometric data.
Other more application-specific chronological methods include shoreline dating [@Brogger1905; @Roalkvam2023], lichenometry [@Benedict2009] and rock weathering dating [@Whitley2012; @Bednarik2020].

@@

Typological dates—i.e. relative, expertise- or seriation-based dating based on artefact characteristics—are ubiquitous in archaeological studies but rarely treated as a form of chronometric data in their own right.
For example, the majority of the radiocarbon datasets we reviewed (@sec-c14-compilation) included some form of typological chronological information in the form of a 'period' or 'culture' column.
This is also typically present in many other forms of systematic compilation work in archaeology, for example site gazetteers.
Aggregated typological information from such sources are often used in aoristic analysis and related methods [@Mischka2004; @Crema2024].
What is lacking in this presentation of typological dating is metadata on how the determination was made and how exactly it is to be understood.
Like any archaeological date, a typological date is derived from a physical sample – the object or set of object from which a chronological estimate was derived.
Typological dates on one class of object may well clash with other classes of object, or for that matter with scientific dates – does one trust the date on pottery, the date on architecture, or the radiocarbon date?
Without additional metadata on e.g. who made the typological determination or what the radiocarbon date was obtained on, such inconsistencies are difficult to resolve.
Similarly the absolute date range corresponding to a typological determination (e.g. "Late Neolithic") can be interpreted in multiple ways depending on the region and intentions of the expert making the determination.
PeriodO [@RabinowitzEtAl2016] is a linked open data infrastructure that includes a shared vocabulary of typological periods and corresponding calendar age estimates, and an important step towards addressing the latter problem.
However, it remains to be systematically linked to compilations of typological dates [though there are some efforts in this direction e.g. @HannahEtAl2022].

@@

The XRONOS data model presents several opportunities to link to other resources as linked open data (LOD).
We use controlled vocabularies such as the GBIF Backbone Taxonomy [@GBIF2023, for taxonomic descriptions of samples] to both standardise fields and to link the two resources based on this shared concept.
There are concordances between XRONOS entities and several other specialised data infrastructures in archaeology, such as PeriodO [@RabinowitzEtAl2016, mapping to typological chrons] and gazetters like Pleiades [@Pleiades] or Vici [@Voorburg2012, mapping to site names].
Moving beyond archaeology, Wikidata (<https://wikidata.org>) already includes many of the concepts represented in the XRONOS model (e.g. an 'archaeological site', <https://www.wikidata.org/wiki/Q839954>).
Linking to Wikidata is especially useful because it dissimenates—and thereby preserves—the data compiled in XRONOS beyond the specialist/academic community.
It also allows us to enrich the database with contextual information that otherwise be beyond our scope and resources, for example embedding multilingual descriptions of sites from Wikipedia articles on site records where the site has been linked to a Wikidata item.

>>>>

Datasets based on literature review also become out of date almost immediately upon publication, due the the constant production of new dates.
Unfortunately this applies to many databases that are in theory continuously updated, as it is common to see them become unmaintained and or unexpectedly become unavailable.
Of the `r n_c14_datasets` published datasets we identified, `r n_c14_datasets_maintained` were intended to be continuously updated, but only `r n_c14_datasets_recently_updated` have received updates in the last two years.
The mean 'lifespan' of a dataset from its publication to its last update is around `r english(round(mean_c14_datasets_lifespan))` years.
Most radiocarbon datasets we reviewed were compiled with a specific goal in mind (e.g. a particular analysis) and, even where there is the intention to keep them updated afterwards, the exigencies of scientific production combined with the labour-intensive nature of the process make that difficult to achieve in practice.

@@

```{r fig-c14-global, message=FALSE, warning=FALSE}
#| fig-cap: Geographic (quantile kernel density) and temporal (sum calibration) of georeferenced dates in XRONOS and other global radiocarbon compilations
#| fig-height: 5
plot_c14_density <- function(density, title, coastline = coast) {
  ggplot() +
    geom_stars(
      aes(fill = ecdf(v)(v)), # transform density to quantiles like plot.stars()
      data = density
    ) +
    geom_sf(data = coastline, fill = NA) +
    scale_fill_stepsn(
      name = "Density",
      n.breaks = 11,
      colours = colour("turku", reverse = TRUE)(11),
      labels = c("Low", rep("", 7), "High"),
      na.value = NA,
      guide = guide_none()
    ) +
    labs(title = title) +
    coord_sf(crs = "+proj=moll") +
    theme_minimal_grid(font_size = 11) +
    theme(
      plot.title = element_text(hjust = 0.5),
      plot.margin = unit(c(11, 5.5, 0, 5.5), "pt")
    )
}

@@

Radiocarbon has been by far the most active area of open data compilation, but archaeological chronology incorporates a much more diverse range of sources of information [@Harding1999].
In periods beyond the practical limit of radiocarbon dating (c. 55,000 BP), other types of radiometric (K–Ar, U–Pb, etc.), chemical or luminescence dating offer an alternative [@Aitken1999].
Conversely, in historic periods, or on calibration 'plateaus' [@GuldersonEtAl2005], radiocarbon is often less important than conventional typological dating (based on artefact characteristics), which in these periods can offer comparable or better temporal resolution, or direct dating based on epigraphy [@HermankovaEtAl2021], numismatics [@KemmersMyrberg2011] or historical sources.
In places where it is widely available, dendrochronology [@Baillie2014] also produces significantly better resolved chronologies and therefore tends to be the main source of chronometric data.
Other more application-specific chronological methods include shoreline dating [@Brogger1905; @Roalkvam2023], lichenometry [@Benedict2009] and rock weathering dating [@Whitley2012; @Bednarik2020; @AndreaeEtAl2020].

@@

Typological dates—i.e. relative, expertise- or seriation-based dating based on artefact characteristics—are ubiquitous in archaeological studies but rarely treated as a form of chronometric data in their own right.
For example, the majority of the radiocarbon datasets we reviewed (@sec-c14-compilation) included some form of typological chronological information in the form of a 'period' or 'culture' column.
This is also typically present in many other forms of systematic compilation work in archaeology, for example site gazetteers.
Aggregated typological information from such sources are often used in aoristic analysis and related methods [@Mischka2004; @Crema2024].
But this presentation of typological dating lacks metadata on how the determination was made and how exactly it is to be understood.
Like any archaeological date, a typological date is derived from a physical sample – the object or set of object from which a chronological estimate was derived.
Typological dates on one class of object may well clash with other classes of object, or for that matter with scientific dates – does one trust the date on pottery, the date on architecture, or the radiocarbon date?
Without additional metadata on e.g. who made the typological determination or what the radiocarbon date was obtained on, such inconsistencies are difficult to resolve.
Similarly the absolute date range corresponding to a typological determination (e.g. "Late Neolithic") can be interpreted in multiple ways depending on the region and intentions of the expert making the determination.
PeriodO [@RabinowitzEtAl2016] is a linked open data infrastructure that includes a shared vocabulary of typological periods and corresponding calendar age estimates, and an important step towards addressing the latter problem.
However, it remains to be systematically linked to compilations of typological dates [though there are some efforts in this direction e.g. @HannahEtAl2022].

@@

The XRONOS data model presents several opportunities to link to other resources as linked open data (LOD).
We use controlled vocabularies such as the GBIF Backbone Taxonomy [@GBIF2023, for taxonomic descriptions of samples] to both standardise fields and to link the two resources based on this shared concept.
There are concordances between XRONOS entities and several other specialised data infrastructures in archaeology, such as PeriodO [@RabinowitzEtAl2016, mapping to typological chrons] and gazetters like Pleiades [@Pleiades] or Vici [@Voorburg2012, mapping to site names].
Moving beyond archaeology, Wikidata (<https://wikidata.org>) already includes many of the concepts represented in the XRONOS model (e.g. an 'archaeological site', <https://www.wikidata.org/wiki/Q839954>).
Linking to Wikidata is especially useful because it disseminates—and thereby preserves—the data compiled in XRONOS beyond the specialist/academic community.
It also allows us to enrich the database with contextual information that otherwise be beyond our scope and resources, for example embedding multilingual descriptions of sites from Wikipedia articles on site records where the site has been linked to a Wikidata item.

### 2025-04-04 paper.tex
Update in response to reviews

Datasets based on literature review also become out of date almost
immediately upon publication, due the the constant production of new
dates. Unfortunately this applies to many databases that are in theory
continuously updated, as it is common to see them become unmaintained
and or unexpectedly become unavailable. Of the 61 published datasets we
identified, 33 were intended to be continuously updated, but only 13
have received updates in the last two years. The average `lifespan' of a
dataset from its publication to its last update is around 4 years. Most
radiocarbon datasets we reviewed were compiled with a specific goal in
mind (e.g.~a particular analysis) and, even where there is the intention
to keep them updated afterwards, the exigencies of scientific production
combined with the labour-intensive nature of the process make that
difficult to achieve in practice.

@@

\caption{\label{fig-c14-global}Geographic (kernel density) and temporal
(sum calibration) of georeferenced dates in XRONOS and other global
radiocarbon compilations}

@@

Radiocarbon has been by far the most active area of open data
compilation, but archaeological chronology incorporates a much more
diverse range of sources of information (Harding, 1999). In periods
beyond the practical limit of radiocarbon dating (c.~55,000 BP), other
types of radiometric (K--Ar, U--Pb, etc.), chemical or luminescence
dating offer an alternative (Aitken, 1999). Conversely, in historic
periods, or on calibration `plateaus' (Guilderson et al., 2005),
radiocarbon is often less important than conventional typological dating
(based on artefact characteristics), which in these periods can offer
comparable or better temporal resolution, or direct dating based on
epigraphy (Heřmánková et al., 2021), numismatics (Kemmers and Myrberg,
2011) or historical sources. In places where it is widely available,
dendrochronology (Baillie, 2014) also produces significantly better
resolved chronologies and therefore tends to be the main source of
chronometric data. Other more application-specific chronological methods
include shoreline dating (Brøgger, 1905; Roalkvam, 2023), lichenometry
(Benedict, 2009) and rock weathering dating (Bednarik, 2020; Whitley,
2012).

@@

Typological dates---i.e.~relative, expertise- or seriation-based dating
based on artefact characteristics---are ubiquitous in archaeological
studies but rarely treated as a form of chronometric data in their own
right. For example, the majority of the radiocarbon datasets we reviewed
(Section~\ref{sec-c14-compilation}) included some form of typological
chronological information in the form of a `period' or `culture' column.
This is also typically present in many other forms of systematic
compilation work in archaeology, for example site gazetteers. Aggregated
typological information from such sources are often used in aoristic
analysis and related methods (Crema, 2024; Mischka, 2004). What is
lacking in this presentation of typological dating is metadata on how
the determination was made and how exactly it is to be understood. Like
any archaeological date, a typological date is derived from a physical
sample -- the object or set of object from which a chronological
estimate was derived. Typological dates on one class of object may well
clash with other classes of object, or for that matter with scientific
dates -- does one trust the date on pottery, the date on architecture,
or the radiocarbon date? Without additional metadata on e.g.~who made
the typological determination or what the radiocarbon date was obtained
on, such inconsistencies are difficult to resolve. Similarly the
absolute date range corresponding to a typological determination
(e.g.~``Late Neolithic'') can be interpreted in multiple ways depending
on the region and intentions of the expert making the determination.
PeriodO (Rabinowitz et al., 2016) is a linked open data infrastructure
that includes a shared vocabulary of typological periods and
corresponding calendar age estimates, and an important step towards
addressing the latter problem. However, it remains to be systematically
linked to compilations of typological dates (though there are some
efforts in this direction e.g. Hannah et al., 2022).

@@

The XRONOS data model presents several opportunities to link to other
resources as linked open data (LOD). We use controlled vocabularies such
as the GBIF Backbone Taxonomy (GBIF Secretariat, 2023, for taxonomic
descriptions of samples) to both standardise fields and to link the two
resources based on this shared concept. There are concordances between
XRONOS entities and several other specialised data infrastructures in
archaeology, such as PeriodO (Rabinowitz et al., 2016, mapping to
typological chrons) and gazetters like Pleiades (Bagnall et al., 2006)
or Vici (Voorburg, 2012, mapping to site names). Moving beyond
archaeology, Wikidata (\url{https://wikidata.org}) already includes many
of the concepts represented in the XRONOS model (e.g.~an `archaeological
site', \url{https://www.wikidata.org/wiki/Q839954}). Linking to Wikidata
is especially useful because it dissimenates---and thereby
preserves---the data compiled in XRONOS beyond the specialist/academic
community. It also allows us to enrich the database with contextual
information that otherwise be beyond our scope and resources, for
example embedding multilingual descriptions of sites from Wikipedia
articles on site records where the site has been linked to a Wikidata
item.

@@

>>>>

Datasets based on literature review also become out of date almost
immediately upon publication, due the the constant production of new
dates. Unfortunately this applies to many databases that are in theory
continuously updated, as it is common to see them become unmaintained
and or unexpectedly become unavailable. Of the 61 published datasets we
identified, 33 were intended to be continuously updated, but only 13
have received updates in the last two years. The mean `lifespan' of a
dataset from its publication to its last update is around 4 years. Most
radiocarbon datasets we reviewed were compiled with a specific goal in
mind (e.g.~a particular analysis) and, even where there is the intention
to keep them updated afterwards, the exigencies of scientific production
combined with the labour-intensive nature of the process make that
difficult to achieve in practice.

@@

\caption{\label{fig-c14-global}Geographic (quantile kernel density) and
temporal (sum calibration) of georeferenced dates in XRONOS and other
global radiocarbon compilations}

@@

Radiocarbon has been by far the most active area of open data
compilation, but archaeological chronology incorporates a much more
diverse range of sources of information (Harding, 1999). In periods
beyond the practical limit of radiocarbon dating (c.~55,000 BP), other
types of radiometric (K--Ar, U--Pb, etc.), chemical or luminescence
dating offer an alternative (Aitken, 1999). Conversely, in historic
periods, or on calibration `plateaus' (Guilderson et al., 2005),
radiocarbon is often less important than conventional typological dating
(based on artefact characteristics), which in these periods can offer
comparable or better temporal resolution, or direct dating based on
epigraphy (Heřmánková et al., 2021), numismatics (Kemmers and Myrberg,
2011) or historical sources. In places where it is widely available,
dendrochronology (Baillie, 2014) also produces significantly better
resolved chronologies and therefore tends to be the main source of
chronometric data. Other more application-specific chronological methods
include shoreline dating (Brøgger, 1905; Roalkvam, 2023), lichenometry
(Benedict, 2009) and rock weathering dating (Andreae et al., 2020;
Bednarik, 2020; Whitley, 2012).

@@

Typological dates---i.e.~relative, expertise- or seriation-based dating
based on artefact characteristics---are ubiquitous in archaeological
studies but rarely treated as a form of chronometric data in their own
right. For example, the majority of the radiocarbon datasets we reviewed
(Section~\ref{sec-c14-compilation}) included some form of typological
chronological information in the form of a `period' or `culture' column.
This is also typically present in many other forms of systematic
compilation work in archaeology, for example site gazetteers. Aggregated
typological information from such sources are often used in aoristic
analysis and related methods (Crema, 2024; Mischka, 2004). But this
presentation of typological dating lacks metadata on how the
determination was made and how exactly it is to be understood. Like any
archaeological date, a typological date is derived from a physical
sample -- the object or set of object from which a chronological
estimate was derived. Typological dates on one class of object may well
clash with other classes of object, or for that matter with scientific
dates -- does one trust the date on pottery, the date on architecture,
or the radiocarbon date? Without additional metadata on e.g.~who made
the typological determination or what the radiocarbon date was obtained
on, such inconsistencies are difficult to resolve. Similarly the
absolute date range corresponding to a typological determination
(e.g.~``Late Neolithic'') can be interpreted in multiple ways depending
on the region and intentions of the expert making the determination.
PeriodO (Rabinowitz et al., 2016) is a linked open data infrastructure
that includes a shared vocabulary of typological periods and
corresponding calendar age estimates, and an important step towards
addressing the latter problem. However, it remains to be systematically
linked to compilations of typological dates (though there are some
efforts in this direction e.g. Hannah et al., 2022).

@@

The XRONOS data model presents several opportunities to link to other
resources as linked open data (LOD). We use controlled vocabularies such
as the GBIF Backbone Taxonomy (GBIF Secretariat, 2023, for taxonomic
descriptions of samples) to both standardise fields and to link the two
resources based on this shared concept. There are concordances between
XRONOS entities and several other specialised data infrastructures in
archaeology, such as PeriodO (Rabinowitz et al., 2016, mapping to
typological chrons) and gazetters like Pleiades (Bagnall et al., 2006)
or Vici (Voorburg, 2012, mapping to site names). Moving beyond
archaeology, Wikidata (\url{https://wikidata.org}) already includes many
of the concepts represented in the XRONOS model (e.g.~an `archaeological
site', \url{https://www.wikidata.org/wiki/Q839954}). Linking to Wikidata
is especially useful because it disseminates---and thereby
preserves---the data compiled in XRONOS beyond the specialist/academic
community. It also allows us to enrich the database with contextual
information that otherwise be beyond our scope and resources, for
example embedding multilingual descriptions of sites from Wikipedia
articles on site records where the site has been linked to a Wikidata
item.

@@

\bibitem[\citeproctext]{ref-AndreaeEtAl2020}
Andreae, M.O., Al-Amri, A., Andreae, C.M., Guagnin, M., Jochum, K.P.,
Stoll, B., Weis, U., 2020. Archaeometric studies on the petroglyphs and
rock varnish at {Kilwa} and {Sakaka}, northern {Saudi Arabia}. Arabian
Archaeology and Epigraphy 31, 219--244.
\url{https://doi.org/10.1111/aae.12167}
